---
title: "Group 8 and AI Knowledge and Willingness to Share Data in Healthcare"
subtitle: "DACSS 602 (Spring 2025)"
author: Ratan Koneru, Jadyn Perrotta, and Mala Mahadevan
format: 
  html:
    toc: true
    toc-depth: 2
    toc-title: Contents
    toc-location: left
    html-math-method: katex
    theme: flatly
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    citations-hover: true
    footnotes-hover: true
    font-size: 80%
---

<!--
ATTENTION: Please submit both your .qmd file and the corresponding ".html" file after rendering.
-->


<!--
Replace the placeholders (title and author) in the yaml above with your own.
-->


# Introduction

Sharing personal information has been a normal part of healthcare for a long time. 
But it is only recently that AI has had the potential to use that information to
help with research, as well as develop individualized treatment plans. These 
opportunities require individuals to trust the system with their data. Our focus
is around studying people’s knowledge/awareness of AI and how that influences their 
willingness to share their personal information with AI in a medical context. Our
hypothesis is that people with more knowledge of AI in healthcare would be more
willing to share their data.



# Methodology
<!--
Describe your experiment design, including: (1) experiment design, (2) treatments, (3) outcome variable(s), (4) sample size, and (5) data cleaning procedures (if applicable) [8-10 sentences]
-->

Our survey experiment was a between subjects design. The treatment group was presented 
with a short infographic on how AI is used in healthcare and the control group  
did not receive any additional information. Both groups then took a survey asking
about their knowledge of AI in healthcare, history of chronic illness, and their 
willingness to share medical data with AI for research and personal treatment. Our 
independent variables were knowledge of AI and chronic illness. Our outcome
variable was willingness to share medical data with AI. We had a sample size of
239 subjects. The survey data was relatively clean - no invalid entries or 'NA' 
responses. Our processing involved adding dummy variables to translate verbal
responses into categorical/continues variables for analysis.


#Hypothesis

We had three hypothesis to test with our data.

1) People shown the infographic on AI (treatment group) will show increased 
willingness to share personal data with AI applications
H0 - The treatment has no effect on willingness to share personal data with AI
H1 - The treatment has significant effect on willingness to share personal data
     with AI
     
2) People with one or more chronic illness will show increased willingness to
share to share personal data with AI applications
H0 - Chronic illness has no effect on willingness to share personal data with AI
H1 - Chronic illness has significant effect on willingness to share personal data
     with AI
     
3) People with self-reported increased knowledge of AI will show increased
willingness to share personal data with AI applications
H0 - Increased knowledge of AI has no effect on willingness to share personal
     data with AI
H1 - Increased knowledge of AI has significant effect on willingness to share
     personal data with AI



# Analysis

<!--
Which statistical test is used, and why? [3-4 sentences]
-->
For our first hypothesis we used a t-test. Assumptions to use t-test:
1 Is the outcome variable continues? It is a categorical variable with 5 categories 
so may be deemed continues for this exercise.
2 Are the test groups independent of one another? Yes
3 Is the outcome normally distributed? Assuming a yes.
4 Are the variances equal between outcomes? 

For our second hypothesis we ran a factorial ANOVA to see whether chronic illness
had an effect together with the treatment/control group on willingness to share
data with AI. We also ran a t-test with just chronic illness to see whether chronic 
illness alone had an effect on willingness to share data with AI.

For our third hypothesis we ran an ANOVA since the independent variable (knowledge
of AI) had more than two levels.



<!--
Run the chosen test! To do so: (1) import the libraries you need, (2) import your data, (3) double-check the names of your IV(s) and DV, and (4) run the test. For information on the libraries and R syntax, refer to see Lecture 602-statistics(May 1).
-->

```{r}
# import libraries
library(readr)
library(tidyverse)
library(tidyr)
library(ggplot2)
library(dplyr)
```


```{r}
#read in data and select variables for analysis
surveydata <- read_csv("Data/SP25_602_omnibus_V2_May8.csv")

filtered_survey <- surveydata |> 
  select(Gender,Age,G8_q2_dv,G8_q3_dv,G8_q4,G8_q5,G8_q6,G8_AI_Health_DO)

#add dummy variables for analysis
filtered_survey <- filtered_survey |> mutate(AIApplications = case_when(
    G8_q2_dv == "5 - Strongly Agree" ~ 5,
    G8_q2_dv == "4 - Agree" ~ 4,
    G8_q2_dv == "3 - Neutral" ~ 3,
    G8_q2_dv == "2 - Disagree" ~ 2,
    G8_q2_dv == "1 - Strongly Disagree" ~ 1
  ))

filtered_survey <- filtered_survey |> mutate(AIResearch = case_when(
    G8_q3_dv == "5 - Strongly Agree" ~ 5,
    G8_q3_dv == "4 - Agree" ~ 4,
    G8_q3_dv == "3 - Neutral" ~ 3,
    G8_q3_dv == "2- Disagree" ~ 2,
    G8_q3_dv == "1 - Strongly Disagree" ~ 1
  ))

filtered_survey <- filtered_survey |> mutate(AIUseinHealthCare = case_when(
    G8_q4 == "Never heard of" ~ 1,
    G8_q4 == "Slightly aware" ~ 2,
    G8_q4 == "Somewhat aware" ~ 3,
    G8_q4 == "Moderately aware" ~ 4,
    G8_q4 == "Extremely aware" ~ 5
  ))

filtered_survey <- filtered_survey |> mutate(AITraining = case_when(
    G8_q5 == "Yes" ~ 1,
    G8_q5 == "No" ~ 0
  ))

filtered_survey <- filtered_survey |> mutate(ChronicIllness = case_when(
    G8_q6 == "None of these" ~ 0,
    G8_q6 != "None of these" ~ 1
  ))

filtered_survey$ControlorTreatment <- ifelse(grepl("treatment", filtered_survey$G8_AI_Health_DO), 1, 0)

filtered_survey <- filtered_survey |>
  mutate(G8_q4 = factor(G8_q4, levels = c("Never heard of", "Slightly aware",
                             "Somewhat aware", "Moderately aware",
                             "Extremely aware"),
                  ordered = TRUE))
#filter data by gender
data_males <- filter(filtered_survey, Gender == "Male")
data_females <- filter(filtered_survey, Gender == "Female")

```

#Tests
```{r}

#Hypothesis 1
t.test(filtered_survey$AIApplications ~ as.factor(filtered_survey$ControlorTreatment), var.equal = FALSE)

#Hypothesis 2
model <- aov(AIApplications ~ as.factor(ControlorTreatment) * ChronicIllness, data = filtered_survey)
summary(model)

t.test(filtered_survey$AIApplications ~ as.factor(filtered_survey$ChronicIllness), var.equal = FALSE)

#Hypothesis 3
model <- aov(filtered_survey$AIApplications ~ as.factor(filtered_survey$AIUseinHealthCare) ,data = filtered_survey)
summary(model)

#Test if gender affects hypothesis 3
model <- aov(AIApplications ~ AIUseinHealthCare ,data = data_males)
summary(model)

model <- aov(AIApplications ~ AIUseinHealthCare ,data = data_females)
summary(model)



```

# Results
<!--
Report and interpret the test results. For information on what should be included, refer to Lecture 602-statistics(May 1). Also add at least one plot.
-->

Test Hypothesis 1:
The p value is 0.1109 , significantly greater than 0.05 which indicates that we cannot reject the null hypothesis. Mean of control group is 2.94 and treatment group is 3.20.
The treatment has not significantly improved outcome - willingness to share data with AI applications.

Test Hypothesis 2:
The p value of the factorial ANOVA is 0.110 indicating lack of statistical evidence for any correlation.
The p value of the t-test is .315 also indicating a lack of statstical evidence for correlation

Test Hypothesis 3:
The low p value of 1.05e-08 suggests statistical significance and we can confidently reject the null hypothesis that knowledge of AI does not significantly impact willingness to share data with AI 
Applications. Low p-values of ANOVAs by gender suggest effect exists in both survey genders.

```{r}
#box plot of willingness to share vs AI knowledge
ggplot(filtered_survey, aes(x = G8_q4, y = AIApplications)) +
  geom_boxplot()

```

# Findings
<!--
Findings: Report the statistic you obtained along with the information to decide whether the null hypothesis can be rejected or not. Use a plot and/or table with a proper label.
-->



# Discussion
<!--
What are the implications of the study? What are possible explanations if the results do not align with your hypothesis? Any suggestions for future studies?
-->



### For your information
<!--
Some example code for different plots. Below is only for your information. Submit your report after removing this section.
-->

# the built-in R data set named ToothGrowth:
# data from a study evaluating the effect of Vitamin C on tooth growth in Guinea pigs.
# The experiment has been performed on 60 pigs, where each animal received one of three dose levels of vitamin C (0.5, 1, and 2 mg/day) by one of two delivery methods, orange juice or ascorbic acid (a form of vitamin C and coded as VC). Tooth length was measured and a sample of the data is shown below
my_data = ToothGrowth
str(my_data)

# Convert dose as a factor and recode the levels
# as D0.5, D1, D2
my_data$dose <- factor(my_data$dose,
                       levels = c(0.5, 1, 2),
                       labels = c("D0.5", "D1", "D2"))
head(my_data)

# frequency table:
table(my_data$supp, my_data$dose)
```


#```{r}
# Line plots with multiple groups
# plot tooth length ("len") by groups ("dose")
# color box plot by a second group: "supp"
# add error bars: mean_se
library(ggpubr)
ggline(my_data, x = "dose", y = "len", color = "supp",
       add = c("mean_se", "dotplot"),
       palette = c("#00AFBB", "#E7B800"))

# OR
# using R base graphs:
boxplot(len ~ supp * dose, data = my_data, frame = FALSE,
        col = c("#00AFBB", "#E7B800"), ylab = "Tooth Length")

# Another example: Two-way interaction plot
interaction.plot(x.factor = my_data$dose, trace.factor = my_data$supp,
                 # x.factor: the factor to be plotted on x axis
                 # trace.factor: the factor to be plotted as lines
                 # response: a numeric variable giving the response
                 response = my_data$len, fun = mean,
                 # type: the type of plot (p: point, l: line, b: both point and line)
                 type = "b", legend = TRUE,
                 xlab = "Dose", ylab = "Tooth Length",
                 pch = c(1, 19), col = c("#00AFBB", "#E7B800"))
```

