---
title: "Group 8 and YOUR PROJECT TITLE"
subtitle: "DACSS 602 (Spring 2025)"
author: Ratan Koneru, Jadyn Perrotta, and Mala Mahadevan
format: 
  html:
    toc: true
    toc-depth: 2
    toc-title: Contents
    toc-location: left
    html-math-method: katex
    theme: flatly
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    citations-hover: true
    footnotes-hover: true
    font-size: 80%
---

<!--
ATTENTION: Please submit both your .qmd file and the corresponding ".html" file after rendering.
-->


<!--
Replace the placeholders (title and author) in the yaml above with your own.
-->


# Introduction
<!--
Sharing personal information has been a normal part of healthcare for a long time. 
But it is only recently that AI has had the potential to use that information to
help with research, as well as develop individualized treatment plans. These 
opportunities require individuals to trust the system with their data. Our focus
is around studying people’s knowledge/awareness of AI and how that influences their 
willingness to share their personal information with AI in a medical context. Our
hypothesis is that people with more knowledge of AI in healthcare would be more
willing to share their data.

-->

# Methodology
<!--
Describe your experiment design, including: (1) experiment design, (2) treatments, (3) outcome variable(s), (4) sample size, and (5) data cleaning procedures (if applicable) [8-10 sentences]

Our survey experiment was a between subjects design. The treatment group was presented 
with a short infographic on how AI is used in healthcare and the control group  
did not receive any additional information. Both groups then took a survey asking
about their knowledge of AI in healthcare, history of chronic illness, and their 
willingness to share medical data with AI for research and personal treatment. Our 
independent variables were knowledge of AI and chronic illness. Our outcome
variable was willingness to share medical data with AI. We had a sample size of
239 subjects.
-->

# Analysis

<!--
Which statistical test is used, and why? [3-4 sentences]
-->

<!--
Run the chosen test! To do so: (1) import the libraries you need, (2) import your data, (3) double-check the names of your IV(s) and DV, and (4) run the test. For information on the libraries and R syntax, refer to see Lecture 602-statistics(May 1).
-->

```{r}
# import libraries
library(readr)
library(tidyverse)
library(tidyr)
library(ggplot2)
library(dplyr)
```


```{r echo=FALSE}
data <- read_csv("SurveyResponses.csv")
head(data)
```
```{r echo=FALSE}

str(data)
summary(data)
```

```{r}

# Clean and format 
data_clean <- data %>%
  # Convert willingness to numeric
  mutate(
    G8_q2_dv = as.numeric(sub(" .*", "", G8_q2_dv)),
    
    # Convert AI knowledge to ordered factor
    G8_q4 = factor(G8_q4, 
                  levels = c("Never heard of", "Slightly aware",
                             "Somewhat aware", "Moderately aware",
                             "Extremely aware"),
                  ordered = TRUE),
    
    # Convert chronic illness to binary factor (Yes/No)
    Chronic_Illness = factor(G8_q5, levels = c("No", "Yes")),
    
  ) %>%
  # Remove rows with missing values
  filter(!is.na(G8_q2_dv), !is.na(G8_q4), !is.na(Chronic_Illness))

# sanity check
str(data_clean)
```

```{r}


# ANOVA to compare willingness across AI knowledge groups
anova_result <- aov(G8_q2_dv ~ G8_q4, data = data_clean)
summary(anova_result)

# t-test to compare willingness across Chronic Illness 
t.test(G8_q2_dv ~ Chronic_Illness, data = data_clean)


```

# Results
<!--
Report and interpret the test results. For information on what should be included, refer to Lecture 602-statistics(May 1). Also add at least one plot.
-->




# Findings
<!--
Findings: Report the statistic you obtained along with the information to decide whether the null hypothesis can be rejected or not. Use a plot and/or table with a proper label.
-->



# Discussion
<!--
What are the implications of the study? What are possible explanations if the results do not align with your hypothesis? Any suggestions for future studies?
-->



### For your information
<!--
Some example code for different plots. Below is only for your information. Submit your report after removing this section.
-->

# the built-in R data set named ToothGrowth:
# data from a study evaluating the effect of Vitamin C on tooth growth in Guinea pigs.
# The experiment has been performed on 60 pigs, where each animal received one of three dose levels of vitamin C (0.5, 1, and 2 mg/day) by one of two delivery methods, orange juice or ascorbic acid (a form of vitamin C and coded as VC). Tooth length was measured and a sample of the data is shown below
my_data = ToothGrowth
str(my_data)

# Convert dose as a factor and recode the levels
# as D0.5, D1, D2
my_data$dose <- factor(my_data$dose,
                       levels = c(0.5, 1, 2),
                       labels = c("D0.5", "D1", "D2"))
head(my_data)

# frequency table:
table(my_data$supp, my_data$dose)
```


```{r}
# Line plots with multiple groups
# plot tooth length ("len") by groups ("dose")
# color box plot by a second group: "supp"
# add error bars: mean_se
library(ggpubr)
ggline(my_data, x = "dose", y = "len", color = "supp",
       add = c("mean_se", "dotplot"),
       palette = c("#00AFBB", "#E7B800"))

# OR
# using R base graphs:
boxplot(len ~ supp * dose, data = my_data, frame = FALSE,
        col = c("#00AFBB", "#E7B800"), ylab = "Tooth Length")

# Another example: Two-way interaction plot
interaction.plot(x.factor = my_data$dose, trace.factor = my_data$supp, 
                 # x.factor: the factor to be plotted on x axis
                 # trace.factor: the factor to be plotted as lines
                 # response: a numeric variable giving the response
                 response = my_data$len, fun = mean,
                 # type: the type of plot (p: point, l: line, b: both point and line)
                 type = "b", legend = TRUE,
                 xlab = "Dose", ylab = "Tooth Length",
                 pch = c(1, 19), col = c("#00AFBB", "#E7B800"))
```

